A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-16 tidyindex.qmd
minor edit and changes in GGGI examples

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations. These dimensions are then combined though another linear combination. @fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index. @tbl-gggi-weights presents the The variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction. In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1. This weights ensures that a one percentage point change in the standard deviation of each variable to contribute equally to the index. Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight. The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`.

@@

The index can be reproduce with tidyindex as:

@@

A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making. We use a visualisation method called "tour" to animate the data projection across different sets of weights, aiming to understand how changes on the politics weight will affect the index value. We select a set countries for this exercise, which includes:

@@

We run a local tour \[TODO: reference to local tour\] to slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions.

@@

In GGGI, the index value has a direct interpretation as the percentage of gender gap that has been closed. When countries exhibit closer index values, or consistent decreasing index values, it can be directly interpreted as country's progress towards gender parity. Analysts must carefully consider whether such interpretation is reasonable and review the index construction process.

@@

```{r fig-idx-tour, fig.align='center', fig.height=5}
#| fig-cap: "Six frames selected from varying the weights of the politics dimension in Global Gender Gap Index (GGGI). The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12. The bottom panel displays the weight used to produce the index value in the top panel, with each frame corresponds to a set of weights. Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6. Compared to Frame 12 where equal weigts are used for the four dimension, a reduced weight in politics (Frame 1 and 6) shows narrower gaps between top and mid- or lower- ranked countries, while an increase in the politics weight (Frame 18, 24, and 29) leads to a systematic decrease of GGGI values across all the countries, except for Bangladesh. Full animation is available at https://vimeo.com/847874016."  
#frames <- c("001", "013", "037", "061", "085", "109")
frames <- c("001", "006", "012", "018", "024", "029")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3)  
```

>>>>

The Global Gender Gap Index (GGGI), published annually by the World Economic Forum, measures gender parity by assessing relative gaps between men and women in four key areas: Economic Participation and Opportunity, Educational Attainment, Health and Survival, and Political Empowerment [@WEF2023]. The index, compiled from 14 variables expressed as female-to-male ratios, first aggregates these variables into four dimensions through linear combinations. These dimensions are then combined though another linear combination to form the index. @fig-pp-gggi illustrates this pipeline construction by applying the dimension reduction module twice on the data to generate the index. @tbl-gggi-weights presents the variable weights (`V-weight`) and dimension weights (`D-weight`) used in the two dimension reduction. In the table, the variable weights are computed as the inverse of the standard deviation of each variable, scaled to sum to 1. These weights ensure that a one percentage point change in the standard deviation of each variable to contribute equally to the index. Dimension weights are equal across the four dimensions and the last column, `weight`, multiples the variable and dimension weights to produce a single set of weight.

@@

The 2023 GGGI data is available from the Global Gender Gap Report 2023 in the country's economy profile and can be accessed in R via the `tidyindex` package as `gggi` and @tbl-gggi-weights as `gggi_weights`. The index can be reproduced with the pacakge as:

@@

A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making. By making slight adjustments to the weight vector, we can observe how index values and country rankings change. For illustration, we select a set countries including:

@@

We slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions. This process creates an animation showing the movement of index values in response to changing weights. This visualisation technique, which presents a sequence of data projections, is referred to as a "tour". The specific kind of tour that moves between the original and nearby projections is known as a "local tour".

@@

In GGGI, the index value has a direct interpretation as the percentage of gender gap that has been closed. When countries exhibit closer index values or consistent, it can be directly interpreted as a reflection of a country's progress toward gender parity. Ideally, an index should be robust against minor changes in its construction components. This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combination. This approach can be applied broadly to different sets of weights of interest, extending beyond the change in a single dimension illustrated by the example.

@@

```{r fig-idx-tour, fig.align='center', fig.height=5}
#| fig-cap: "Six frames selected to explore how varying the weights of the politics dimension changes the index values and country ranking in Global Gender Gap Index (GGGI). The top panel shows the GGGI value against the country, ranked by its original index value in Frame 12. The bottom panel displays the weight used to produce the index value in the top panel, with each frame corresponds to a set of weights. Countries selected includes 1) top-ranked countries with GGGI > 0.85, 2) countries ranked between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked countries with GGGI < 0.6. Compared to Frame 12 where equal weigts are used for the four dimension, a reduced weight in politics (Frame 1 and 6) shows narrower gaps between top and mid- or lower- ranked countries, while an increase in the politics weight (Frame 18, 24, and 29) leads to a systematic decrease of GGGI values across all the countries, except for Bangladesh. Full animation is available at https://vimeo.com/847874016."  
#frames <- c("001", "013", "037", "061", "085", "109")
frames <- c("001", "006", "012", "018", "024", "029")
ani <- paste0(here::here("figures/"), "idx-tour/", "idx-tour-", frames, ".png")
rl <- lapply(ani, png::readPNG)
gl <-  lapply(rl, grid::rasterGrob)
wrap_plots(gl, ncol = 3)  
```

### 2023-08-16 tidyindex.tex
minor edit and changes in GGGI examples

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, sharp corners, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

The Global Gender Gap Index (GGGI), published annually by the World
Economic Forum, measures gender parity by assessing relative gaps
between men and women in four key areas: Economic Participation and
Opportunity, Educational Attainment, Health and Survival, and Political
Empowerment (World Economic Forum 2023). The index, compiled from 14
variables expressed as female-to-male ratios, first aggregates these
variables into four dimensions through linear combinations. These
dimensions are then combined though another linear combination.
Figure~\ref{fig-pp-gggi} illustrates this pipeline construction by
applying the dimension reduction module twice on the data to generate
the index. Table~\ref{tbl-gggi-weights} presents the The variable
weights (\texttt{V-weight}) and dimension weights (\texttt{D-weight})
used in the two dimension reduction. In the table, the variable weights
are computed as the inverse of the standard deviation of each variable,
scaled to sum to 1. This weights ensures that a one percentage point
change in the standard deviation of each variable to contribute equally
to the index. Dimension weights are equal across the four dimensions and
the last column, \texttt{weight}, multiples the variable and dimension
weights to produce a single set of weight. The 2023 GGGI data is
available from the Global Gender Gap Report 2023 in the country's
economy profile and can be accessed in R via the \texttt{tidyindex}
package as \texttt{gggi} and Table~\ref{tbl-gggi-weights} as
\texttt{gggi\_weights}.

@@

The index can be reproduce with tidyindex as:

@@

A linear combination can also be interpreted as a linear projection of
multivariate information with a weight vector. Projecting data from
higher to lower dimension unavoidably leads to information loss and the
weights used require careful examination to understand their effects on
the index and implications for interpretation and decision-making. We
use a visualisation method called ``tour'' to animate the data
projection across different sets of weights, aiming to understand how
changes on the politics weight will affect the index value. We select a
set countries for this exercise, which includes:

@@

We run a local tour {[}TODO: reference to local tour{]} to slightly vary
the weight of the politics dimension from the original 0.25 while
keeping the weights constant for the other three dimensions.

@@

\caption{\label{fig-idx-tour}Six frames selected from varying the
weights of the politics dimension in Global Gender Gap Index (GGGI). The
top panel shows the GGGI value against the country, ranked by its
original index value in Frame 12. The bottom panel displays the weight
used to produce the index value in the top panel, with each frame
corresponds to a set of weights. Countries selected includes 1)
top-ranked countries with GGGI \textgreater{} 0.85, 2) countries ranked
between 57 and 62 with GGGI from 0.72 to 0.73, and 3) low-ranked
countries with GGGI \textless{} 0.6. Compared to Frame 12 where equal
weigts are used for the four dimension, a reduced weight in politics
(Frame 1 and 6) shows narrower gaps between top and mid- or lower-
ranked countries, while an increase in the politics weight (Frame 18,
24, and 29) leads to a systematic decrease of GGGI values across all the
countries, except for Bangladesh. Full animation is available at
https://vimeo.com/847874016.}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

The Global Gender Gap Index (GGGI), published annually by the World
Economic Forum, measures gender parity by assessing relative gaps
between men and women in four key areas: Economic Participation and
Opportunity, Educational Attainment, Health and Survival, and Political
Empowerment (World Economic Forum 2023). The index, compiled from 14
variables expressed as female-to-male ratios, first aggregates these
variables into four dimensions through linear combinations. These
dimensions are then combined though another linear combination to form
the index. Figure~\ref{fig-pp-gggi} illustrates this pipeline
construction by applying the dimension reduction module twice on the
data to generate the index. Table~\ref{tbl-gggi-weights} presents the
variable weights (\texttt{V-weight}) and dimension weights
(\texttt{D-weight}) used in the two dimension reduction. In the table,
the variable weights are computed as the inverse of the standard
deviation of each variable, scaled to sum to 1. These weights ensure
that a one percentage point change in the standard deviation of each
variable to contribute equally to the index. Dimension weights are equal
across the four dimensions and the last column, \texttt{weight},
multiples the variable and dimension weights to produce a single set of
weight.

@@

The 2023 GGGI data is available from the Global Gender Gap Report 2023
in the country's economy profile and can be accessed in R via the
\texttt{tidyindex} package as \texttt{gggi} and
Table~\ref{tbl-gggi-weights} as \texttt{gggi\_weights}. The index can be
reproduced with the pacakge as:

@@

A linear combination can also be interpreted as a linear projection of
multivariate information with a weight vector. Projecting data from
higher to lower dimension unavoidably leads to information loss and the
weights used require careful examination to understand their effects on
the index and implications for interpretation and decision-making. By
making slight adjustments to the weight vector, we can observe how index
values and country rankings change. For illustration, we select a set
countries including:

@@

We slightly vary the weight of the politics dimension from the original
0.25 while keeping the weights constant for the other three dimensions.
This process creates an animation showing the movement of index values
in response to changing weights. This visualisation technique, which
presents a sequence of data projections, is referred to as a ``tour''.
The specific kind of tour that moves between the original and nearby
projections is known as a ``local tour''.

@@

\caption{\label{fig-idx-tour}Six frames selected to explore how varying
the weights of the politics dimension changes the index values and
country ranking in Global Gender Gap Index (GGGI). The top panel shows
the GGGI value against the country, ranked by its original index value
in Frame 12. The bottom panel displays the weight used to produce the
index value in the top panel, with each frame corresponds to a set of
weights. Countries selected includes 1) top-ranked countries with GGGI
\textgreater{} 0.85, 2) countries ranked between 57 and 62 with GGGI
from 0.72 to 0.73, and 3) low-ranked countries with GGGI \textless{}
0.6. Compared to Frame 12 where equal weigts are used for the four
dimension, a reduced weight in politics (Frame 1 and 6) shows narrower
gaps between top and mid- or lower- ranked countries, while an increase
in the politics weight (Frame 18, 24, and 29) leads to a systematic
decrease of GGGI values across all the countries, except for Bangladesh.
Full animation is available at https://vimeo.com/847874016.}
